---
title: "CG3CG3"
author: 
- de Pedro, Joenell Conrad
- Florentino, Ian Miguel
- Nacu, Drixyl Reece
- Ruelan, Ivan Paul
- Seno, Matthew Anakin
date: "CG3CG3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(car)
library(tidyverse)
library(cluster)
library(ggplot2)
```

```{r}
# data <- read.csv("developer_survey.csv")
data <- read.csv("./Culminating_Activity/developer_survey.csv")
head(data)
```

# Load and Inspect Data
Apparently you don't have to convert categorical into numerical 
values in R, it just automatically labels them with the as.factor function

```{r}
sum(is.na(data$ai_tools_used))
head(data)
print("GLIMPSE")
glimpse(data)
print("SUMMARY")
summary(data)
```

# Preprocessing Data

```{r}
data_clean <- data %>%
  
  select(-respondent_id, -languages_used, -frameworks_used,
         -databases_used, -cloud_platforms_used, -ide_editor,
         -want_to_learn_next, -ai_tools_used) %>%   # ai_tools_used has ~20% NAs
  
  mutate(
    education_level = factor(education_level,
      levels = c("Less than high school", "Some high school",
                 "High school diploma", "Some college",
                 "Associate's degree", "Bachelor's degree",
                 "Master's degree", "Professional degree",
                 "Doctoral degree"),
      ordered = TRUE),
    
    company_size = factor(company_size,
      levels = c("1-10", "11-50", "51-200", "201-1000",
                 "1001-5000", "5000-10000", "10000+"),
      ordered = TRUE),
    
    burnout_frequency = factor(burnout_frequency,
      levels = c("Never", "Rarely", "Sometimes", "Often", "Always"),
      ordered = TRUE),
    
    ai_usage_frequency = factor(ai_usage_frequency,
      levels = c("Never", "Rarely", "Monthly", "Weekly",
                 "Daily", "Multiple times daily"),
      ordered = TRUE),
    
    gender              = factor(gender),
    role                = factor(role),
    employment_status   = factor(employment_status),
    work_mode           = factor(work_mode),
    primary_os          = factor(primary_os),
    ai_sentiment        = factor(ai_sentiment),
    ai_job_threat_perception = factor(ai_job_threat_perception),
    primary_learning_source  = factor(primary_learning_source),
    looking_for_job     = factor(looking_for_job),
    country             = factor(country)
  ) %>%
  
  filter(annual_salary_usd > 0) %>%
  
  mutate(log_salary = log(annual_salary_usd))

cat("Rows after cleaning:", nrow(data_clean), "\n")
```

# Explaratory Data Analysis
```{r}
par(mfrow = c(1, 2))
hist(data_clean$annual_salary_usd, main = "Raw Salary", col = "steelblue")
hist(data_clean$log_salary,        main = "Log Salary", col = "tomato")

# Correlation of numeric predictors vs log_salary
num_cols <- data_clean %>%
  select(where(is.numeric)) %>%
  select(-annual_salary_usd, -survey_year)

cor(num_cols, use = "complete.obs") %>%
  as.data.frame() %>%
  select(log_salary) %>%
  arrange(desc(abs(log_salary)))
```

# Model Fitting
```{r}
model <- lm(
  log_salary ~
    years_of_experience +
    age +
    coding_hours_per_week +
    education_level +
    role +
    employment_status +
    company_size +
    work_mode +
    gender +
    primary_os +
    ai_productivity_boost_pct +
    ai_usage_frequency +
    burnout_frequency +
    remote_work_preference_1to5 +
    country +
    open_source_contributor,
  data = data_clean
)
```

# Model Summary
```{r}
summary(model)
anova(model)

tidy_model <- broom::tidy(model)

tidy_model %>%
  filter(p.value < 0.05) %>%
  arrange(p.value) %>%
  print(n = 50)

tidy_model %>%
  filter(p.value >= 0.05) %>%
  arrange(desc(p.value))
```


# Model Versions
## Model v2: Remove non-significant predictors
```{r}
model_v2 <- lm(
  log_salary ~
    years_of_experience +
    employment_status +
    role +
    coding_hours_per_week +
    age +
    education_level,
  data = data_clean
)

summary(model_v2)
anova(model_v2)
```

## Model v3: Add country 
```{r}
model_v3 <- lm(
  log_salary ~
    years_of_experience +
    employment_status +
    role +
    coding_hours_per_week +
    age +
    education_level +
    country,          
  data = data_clean
)

summary(model_v3)
anova(model_v3)
```

## Model v4: Dropped Coding Hours per Week
- Since it is not significant in v3, we will drop it in v4
```{r}  
model_v4 <- lm(
  log_salary ~
    years_of_experience +
    employment_status +
    role +
    coding_hours_per_week +
    age +
    education_level +
    country,          
  data = data_clean
)

summary(model_v4)
anova(model_v4)

```

## Cross Validation Testing
```{r}
library(caret)

set.seed(33)
cv_control <- trainControl(method = "cv", number = 10)  # 10-fold CV

model_cv <- train(
  log_salary ~
    years_of_experience +
    employment_status +
    role +
    age +
    education_level +
    country,
  data    = data_clean,
  method  = "lm",
  trControl = cv_control,
  na.action = na.omit
)

print(model_cv)
model_cv$results   # compare CV R² vs your in-sample R² of 0.9229
```

# Model Diagnostics
```{r}
par(mfrow = c(2, 2))
plot(model)
```
