---
title: "CG3CG3"
author: 
- de Pedro, Joenell Conrad
- Florentino, Ian Miguel
- Nacu, Drixyl Reece
- Ruelan, Ivan Paul
- Seno, Matthew Anakin
date: "CG3CG3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(cluster)
library(car)
library(broom)
library(MASS)
library(leaps)

# Pin select to dplyr — prevents MASS from masking it
select <- dplyr::select
```

```{r load_data}
data <- read.csv("./src/Culminating_Activity/developer_survey.csv")
head(data)
```

# Load and Inspect Data

```{r inspect}
sum(is.na(data$ai_tools_used))
head(data)
print("GLIMPSE")
glimpse(data)
print("SUMMARY")
summary(data)
```

# Preprocessing Data

```{r missing_summary}
missing_summary <- data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(pct_missing = round(n_missing / nrow(data) * 100, 1)) %>%
  filter(n_missing > 0) %>%
  arrange(desc(pct_missing))

print(missing_summary)
```

```{r data_clean}
data_clean <- data %>%

  dplyr::select(-respondent_id, -languages_used, -frameworks_used,
                -databases_used, -cloud_platforms_used, -ide_editor,
                -want_to_learn_next, -ai_tools_used) %>%

  # Impute with median for numeric columns (central tendency)
  mutate(
    years_of_experience         = ifelse(is.na(years_of_experience),
                                    median(years_of_experience, na.rm = TRUE),
                                    years_of_experience),
    age                         = ifelse(is.na(age),
                                    median(age, na.rm = TRUE), age),
    coding_hours_per_week       = ifelse(is.na(coding_hours_per_week),
                                    median(coding_hours_per_week, na.rm = TRUE),
                                    coding_hours_per_week),
    ai_productivity_boost_pct   = ifelse(is.na(ai_productivity_boost_pct),
                                    median(ai_productivity_boost_pct, na.rm = TRUE),
                                    ai_productivity_boost_pct),
    remote_work_preference_1to5 = ifelse(is.na(remote_work_preference_1to5),
                                    median(remote_work_preference_1to5, na.rm = TRUE),
                                    remote_work_preference_1to5)
  ) %>%

  # Impute with mode for categorical columns (most common category)
  mutate(
    across(
      c(gender, role, employment_status, work_mode, primary_os,
        ai_sentiment, ai_job_threat_perception, primary_learning_source,
        looking_for_job, country, open_source_contributor),
      ~ {
        mode_val <- names(sort(table(.), decreasing = TRUE))[1]
        ifelse(is.na(.), mode_val, .)
      }
    )
  ) %>%

  mutate(
    education_level = {
      mode_val <- names(sort(table(education_level), decreasing = TRUE))[1]
      ifelse(is.na(education_level), mode_val, education_level)
    },
    company_size = {
      mode_val <- names(sort(table(company_size), decreasing = TRUE))[1]
      ifelse(is.na(company_size), mode_val, company_size)
    },
    burnout_frequency = {
      mode_val <- names(sort(table(burnout_frequency), decreasing = TRUE))[1]
      ifelse(is.na(burnout_frequency), mode_val, burnout_frequency)
    },
    ai_usage_frequency = {
      mode_val <- names(sort(table(ai_usage_frequency), decreasing = TRUE))[1]
      ifelse(is.na(ai_usage_frequency), mode_val, ai_usage_frequency)
    }
  ) %>%

  mutate(

    education_level = factor(education_level,
      levels = c("High school",
                 "Some college",
                 "Associate degree",
                 "Self-taught / Bootcamp",
                 "Bachelor's degree",
                 "Master's degree",
                 "Doctorate (PhD)"),
      ordered = TRUE),

    # CSV values: "1-10", "11-50", "51-200", "201-1000",
    #             "1001-5000", "5001-10000", "10000+"
    company_size = factor(company_size,
      levels = c("1-10", "11-50", "51-200", "201-1000",
                 "1001-5000", "5001-10000", "10000+"),
      ordered = TRUE),

    burnout_frequency = factor(burnout_frequency,
      levels = c("Never", "Rarely", "Sometimes", "Often", "Always"),
      ordered = TRUE),

    ai_usage_frequency = factor(ai_usage_frequency,
      levels = c("Never", "Rarely", "Monthly", "Weekly",
                 "Daily", "Multiple times daily"),
      ordered = TRUE),

    gender                   = factor(gender),
    role                     = factor(role),
    employment_status        = factor(employment_status),
    work_mode                = factor(work_mode),
    primary_os               = factor(primary_os),
    ai_sentiment             = factor(ai_sentiment),
    ai_job_threat_perception = factor(ai_job_threat_perception),
    primary_learning_source  = factor(primary_learning_source),
    looking_for_job          = factor(looking_for_job),
    country                  = factor(country)
  ) %>%

  filter(!is.na(annual_salary_usd), annual_salary_usd > 0) %>%
  mutate(log_salary = log(annual_salary_usd))
```

```{r verify_clean}
cat("Rows after cleaning:", nrow(data_clean), "\n")

remaining_na <- data_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  filter(n_missing > 0)

if (nrow(remaining_na) == 0) {
  cat("No missing values remaining.\n")
} else {
  cat("WARNING: Missing values still present:\n")
  print(remaining_na)
}
```

# Exploratory Data Analysis

```{r eda}
par(mfrow = c(1, 2))
hist(data_clean$annual_salary_usd, main = "Raw Salary", col = "steelblue")
hist(data_clean$log_salary,        main = "Log Salary", col = "tomato")

# Correlation of numeric predictors vs log_salary
num_cols <- data_clean %>%
  dplyr::select(where(is.numeric)) %>%
  dplyr::select(-annual_salary_usd, -survey_year)

cor(num_cols, use = "complete.obs") %>%
  as.data.frame() %>%
  dplyr::select(log_salary) %>%
  arrange(desc(abs(log_salary)))
```

# Model Fitting

```{r model_base}
model <- lm(
  log_salary ~
    years_of_experience +
    age +
    coding_hours_per_week +
    education_level +
    role +
    employment_status +
    company_size +
    work_mode +
    gender +
    primary_os +
    ai_productivity_boost_pct +
    ai_usage_frequency +
    burnout_frequency +
    remote_work_preference_1to5 +
    country +
    open_source_contributor,
  data = data_clean
)
```

# Model Summary

```{r model_summary}
summary(model)
anova(model)

tidy_model <- broom::tidy(model)

cat("\n--- Significant predictors (p < 0.05) ---\n")
tidy_model %>%
  filter(p.value < 0.05) %>%
  arrange(p.value) %>%
  print(n = 50)

cat("\n--- Non-significant predictors (p >= 0.05) ---\n")
tidy_model %>%
  filter(p.value >= 0.05) %>%
  arrange(desc(p.value))
```

# Model Versions (Manual)

## Model v2: Remove non-significant predictors

```{r model_v2}
model_v2 <- lm(
  log_salary ~
    years_of_experience +
    employment_status +
    role +
    coding_hours_per_week +
    age +
    education_level,
  data = data_clean
)

summary(model_v2)
anova(model_v2)
```

## Model v3: Add country

```{r model_v3}
model_v3 <- lm(
  log_salary ~
    years_of_experience +
    employment_status +
    role +
    coding_hours_per_week +
    age +
    education_level +
    country,
  data = data_clean
)

summary(model_v3)
anova(model_v3)
```

## Model v4: Drop coding_hours_per_week (not significant in v3)

```{r model_v4}
model_v4 <- lm(
  log_salary ~
    years_of_experience +
    employment_status +
    role +
    age +
    education_level +
    country,
  data = data_clean
)

summary(model_v4)
anova(model_v4)
```

# Variable Selection

## Full Model (baseline for selection)

```{r vs_full}
model_full <- lm(
  log_salary ~ years_of_experience + age + coding_hours_per_week +
    education_level + role + employment_status + company_size +
    work_mode + gender + primary_os + ai_productivity_boost_pct +
    ai_usage_frequency + burnout_frequency + remote_work_preference_1to5 +
    country + open_source_contributor,
  data = data_clean
)

summary(model_full)
AIC(model_full)
BIC(model_full)
```

## Backward Elimination by p-value

```{r vs_backward}
drop1(model_full, test = "F")

# Adjust variables below based on YOUR actual drop1 output (drop where p > 0.05)
model_backward_pval2 <- update(model_full, . ~ . - gender - primary_os - burnout_frequency)
summary(model_backward_pval2)

drop1(model_backward_pval2, test = "F")
```

## Stepwise AIC (both directions)

```{r vs_aic}
model_step_aic <- stepAIC(model_full, direction = "both", trace = TRUE)
summary(model_step_aic)
AIC(model_step_aic)
BIC(model_step_aic)
```

## Stepwise BIC (penalizes complexity more)

```{r vs_bic}
n <- nrow(data_clean)
model_step_bic <- stepAIC(model_full, direction = "both", k = log(n), trace = TRUE)
summary(model_step_bic)
AIC(model_step_bic)
BIC(model_step_bic)
```

## Theory-Driven Reduced Model

```{r vs_theory}
model_theory <- lm(
  log_salary ~ years_of_experience + education_level + role +
    employment_status + company_size + work_mode + country,
  data = data_clean
)

summary(model_theory)
AIC(model_theory)
BIC(model_theory)
```

## VIF Check — Remove High Multicollinearity (VIF > 5)

```{r vs_vif}
vif(model_full)

model_no_collinear <- lm(
  log_salary ~ years_of_experience + coding_hours_per_week +
    education_level + role + employment_status + company_size +
    work_mode + gender + primary_os + ai_productivity_boost_pct +
    ai_usage_frequency + burnout_frequency + remote_work_preference_1to5 +
    country + open_source_contributor,
  data = data_clean
)

vif(model_no_collinear)
summary(model_no_collinear)
```

## Model Comparison Table (All Models)

```{r model_comparison}
models <- list(
  Full            = model_full,
  Manual_v2       = model_v2,
  Manual_v3       = model_v3,
  Manual_v4       = model_v4,
  Backward_Pval   = model_backward_pval2,
  Stepwise_AIC    = model_step_aic,
  Stepwise_BIC    = model_step_bic,
  Theory_Driven   = model_theory,
  No_Collinear    = model_no_collinear
)

comparison <- data.frame(
  Model    = names(models),
  R2       = sapply(models, function(m) round(summary(m)$r.squared, 4)),
  Adj_R2   = sapply(models, function(m) round(summary(m)$adj.r.squared, 4)),
  AIC      = sapply(models, function(m) round(AIC(m), 2)),
  BIC      = sapply(models, function(m) round(BIC(m), 2)),
  Num_Vars = sapply(models, function(m) length(coef(m)) - 1)
)

comparison <- comparison[order(comparison$Adj_R2, decreasing = TRUE), ]
print(comparison)

cat("\nBest model by Adj R2:", comparison$Model[1], "\n")
cat("Best model by AIC:   ", comparison$Model[which.min(comparison$AIC)], "\n")
cat("Best model by BIC:   ", comparison$Model[which.min(comparison$BIC)], "\n")
```

# Final Model Summary
```{r}
final_model <- model_full
summary(final_model)
anova(final_model)
```

# Diagnostic Plots for Final Model
## Residual Plots
```{r}

par(mfrow = c(2, 2))
plot(final_model)
par(mfrow = c(1, 1))
```
## Q-Q Plot
```{r}
par(mfrow = c(1, 2))

qqnorm(residuals(final_model), main = "Normal Q-Q Plot of Residuals")
qqline(residuals(final_model), col = "red", lwd = 2)

# Visual: Histogram of residuals
hist(residuals(final_model),
     breaks = 50,
     col    = "steelblue",
     main   = "Histogram of Residuals",
     xlab   = "Residuals")
curve(dnorm(x,
            mean = mean(residuals(final_model)),
            sd   = sd(residuals(final_model))) * length(residuals(final_model)) *
        diff(hist(residuals(final_model), breaks = 50, plot = FALSE)$breaks)[1],
      add = TRUE, col = "red", lwd = 2)

par(mfrow = c(1, 1))

```
### Shapiro Walk
```{r}
set.seed(33)
resid_sample <- sample(residuals(final_model), min(5000, length(residuals(final_model))))
shapiro_result <- shapiro.test(resid_sample)
print(shapiro_result)
cat("Shapiro-Wilk: W =", round(shapiro_result$statistic, 4),
    " | p-value =", shapiro_result$p.value, "\n")
cat("Normality assumption:", ifelse(shapiro_result$p.value > 0.05, "PASSED", "VIOLATED"), "\n\n")

ks_result <- ks.test(residuals(final_model), "pnorm",
                     mean = mean(residuals(final_model)),
                     sd   = sd(residuals(final_model)))
print(ks_result)
cat("KS Test: D =", round(ks_result$statistic, 4),
    " | p-value =", ks_result$p.value, "\n")
cat("Normality assumption:", ifelse(ks_result$p.value > 0.05, "PASSED", "VIOLATED"), "\n\n")
```
## HOMOSCEDASTICITY 
### Residual vs Fitted
```{r}
plot(fitted(final_model), residuals(final_model),
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals",
     pch  = 20, col = rgb(0, 0, 1, 0.3))
abline(h = 0, col = "red", lwd = 2)
lines(lowess(fitted(final_model), residuals(final_model)), col = "orange", lwd = 2)
```

### Scale-location Plot
```{r}
plot(fitted(final_model), sqrt(abs(residuals(final_model))),
     main = "Scale-Location",
     xlab = "Fitted Values",
     ylab = expression(sqrt("|Residuals|")),
     pch  = 20, col = rgb(0, 0, 1, 0.3))
lines(lowess(fitted(final_model), sqrt(abs(residuals(final_model)))),
      col = "red", lwd = 2)
```
### BPtest
```{r}
if (!require(lmtest)) install.packages("lmtest")
library(lmtest)

bp_result <- bptest(final_model)
print(bp_result)
cat("Breusch-Pagan: BP =", round(bp_result$statistic, 4),
    " | p-value =", bp_result$p.value, "\n")
cat("Homoscedasticity assumption:", ifelse(bp_result$p.value > 0.05, "PASSED", "VIOLATED"), "\n\n")
```

## Multicollinearity Check
```{r}
vif_vals <- vif(final_model)
print(vif_vals)

cat("\n--- VIF > 5 (moderate concern) ---\n")
print(vif_vals[vif_vals > 5])

cat("\n--- VIF > 10 (severe concern) ---\n")
print(vif_vals[vif_vals > 10])

cat("Multicollinearity assumption:", 
    ifelse(all(vif_vals < 10), "PASSED (no VIF > 10)", "VIOLATED"), "\n\n")
```

## Independence Check (Durbin-Watson Test)
```{r}

dw_result <- dwtest(final_model)
print(dw_result)
cat("Durbin-Watson: DW =", round(dw_result$statistic, 4),
    " | p-value =", dw_result$p.value, "\n")
cat("Independence assumption:", 
    ifelse(dw_result$statistic > 1.5 & dw_result$statistic < 2.5,
           "PASSED", "CHECK — DW outside 1.5–2.5 range"), "\n\n")

```
## Cooks Distance
```{r}
cooksd    <- cooks.distance(final_model)
threshold <- 4 / nrow(data_clean)   

influential <- which(cooksd > threshold)
cat("Number of influential points (Cook's D > 4/n):", length(influential), "\n")
cat("That's", round(length(influential) / nrow(data_clean) * 100, 1), "% of observations\n\n")

# Visual
plot(cooksd, type = "h",
     main = "Cook's Distance",
     ylab = "Cook's Distance",
     xlab = "Observation Index",
     col  = ifelse(cooksd > threshold, "red", "steelblue"))
abline(h = threshold, col = "red", lty = 2)
legend("topright", legend = paste0("Threshold = 4/n = ", round(threshold, 5)),
       col = "red", lty = 2)

cat("Top 10 most influential observations:\n")
top_influential <- sort(cooksd, decreasing = TRUE)[1:10]
print(round(top_influential, 6))
```

## Leverage (Hat Values)
```{r}
hat_vals  <- hatvalues(final_model)
p         <- length(coef(final_model))     # number of parameters
n         <- nrow(data_clean)
lev_threshold <- 2 * p / n                # common rule: 2p/n

high_leverage <- which(hat_vals > lev_threshold)
cat("\nHigh leverage points (hat > 2p/n =", round(lev_threshold, 4), "):",
    length(high_leverage), "\n")
```

## Summary
```{r}
cat("\n============================================================\n")
cat("DIAGNOSTIC SUMMARY\n")
cat("============================================================\n")
cat(sprintf("%-30s %-10s %s\n", "Test", "Result", "Status"))
cat(sprintf("%-30s %-10s %s\n", "----", "------", "------"))
cat(sprintf("%-30s %-10s %s\n", "Shapiro-Wilk (normality)",
    round(shapiro_result$p.value, 4),
    ifelse(shapiro_result$p.value > 0.05, "PASSED", "VIOLATED")))
cat(sprintf("%-30s %-10s %s\n", "KS Test (normality)",
    round(ks_result$p.value, 4),
    ifelse(ks_result$p.value > 0.05, "PASSED", "VIOLATED")))
cat(sprintf("%-30s %-10s %s\n", "Breusch-Pagan (homosced.)",
    round(bp_result$p.value, 4),
    ifelse(bp_result$p.value > 0.05, "PASSED", "VIOLATED")))
cat(sprintf("%-30s %-10s %s\n", "Durbin-Watson (independence)",
    round(dw_result$statistic, 4),
    ifelse(dw_result$statistic > 1.5 & dw_result$statistic < 2.5, "PASSED", "CHECK")))
cat(sprintf("%-30s %-10s %s\n", "VIF max (multicollinearity)",
    round(max(vif_vals), 4),
    ifelse(max(vif_vals) < 10, "PASSED", "VIOLATED")))
cat(sprintf("%-30s %-10s %s\n", "Influential points (Cook's D)",
    length(influential),
    ifelse(length(influential) / n < 0.05, "ACCEPTABLE", "REVIEW")))
cat("============================================================\n")
```

