---
title: "CG3CG3"
author: 
- de Pedro, Joenell Conrad
- Florentino, Ian Miguel
- Nacu, Drixyl Reece
- Ruelan, Ivan Paul
- Seno, Matthew Anakin
date: "CG3CG3"
output: "html_document"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# 1. Clear environment (optional but recommended to prevent ghost variables)
rm(list = ls())

# 2. Set seed and define n
set.seed(33)
n <- 100

# 3. Generate the dataframe
data <- data.frame(
  StudyHours = round(runif(n, 2, 20), 1),
  AttendanceRate = round(runif(n, 60, 100), 1),
  Gender = factor(sample(c("Male", "Female"), n, replace = TRUE)),
  SchoolType = factor(sample(c("Public", "Private"), n, replace = TRUE))
)

# 4. Calculate ScienceScore
data$ScienceScore <- round(
  40 +
    2.5 * data$StudyHours +
    0.3 * data$AttendanceRate +
    ifelse(data$Gender == "Female", 3, 0) +
    ifelse(data$SchoolType == "Private", 5, 0) +
    rnorm(n, 0, 5), 1
)

head(data)
model <- lm(ScienceScore ~ ., data = data)

# View the coefficients (Intercept and Slope) to form your equation
coef(model)

summary(model)
```
# Normality
## QQplot
```{r}
qqnorm (residuals(model), ylab="Residuals")
qqline(residuals(model))
```
## Shapiro Test
```{r}
length(residuals(model))
shapiro.test(residuals(model))
```
- Does not reject the null hypothesis

## Statistical Test
```{r}
resid_std <- scale(residuals(model)) # mean=0, sd=1
ks.test(resid_std, "pnorm")
```
# Homoscedasticity

## Graphical Analysis of the residuals vs fitted values
```{r}
plot(fitted(model), residuals(model))
```

## Statistical Test
```{r}
library("lmtest")
bptest(model)
```
- Homoscedasticity is not present in the model since the p-value is greater than 0.05

## Goldfeld Quandt Test
```{r}
gqtest(model, order.by = ScienceScore ~ ., data = data)
```
- GQ value > 1 suggests variance might be larger in the second segment

- p-value = 0.0611 which is greater than 0.05 suggests that there is no heteroscedasticity which also makes it borderline significant


# Multicollinearity
PURPOSE OF MULTICOLLINEARITY: Check if the features are not highly correlated with each other
```{r}
options(width = 50)
library(car)
vif(model)
# VIF > 5 -> moderate concern
# VIF > 10 -> serious multicollinearity
```
Since there is no high correlation with other independent variables, so it does not have a multicollinearity issue.

# Independence Test

## Graphical Analysis of Residuals
```{r}
n = length(residuals(mmodel))
plot(tail(residuals(mmodel), n-1) ~
head(residuals(mmodel), n-1),
xlab=expression(hat(epsilon)[i]),
ylab=expression(hat(epsilon)[i+1]))
abline(h=0, v=0, col = grey(0.5))
```

## Statistical Test
```{r}
dwtest(model)
```
- The p-value indicates no evidence of serial correlation. Therefore, we can assume that the errors are uncorrelated

# Regression Diagnostics Plot
```{r}
par(mfrow = c(2, 2))
plot(model)
```
- Here are the diagnostic plots for each regression