---
title: "CG3CG3"
author: 
- de Pedro, Joenell Conrad
- Florentino, Ian Miguel
- Nacu, Drixyl Reece
- Ruelan, Ivan Paul
- Seno, Matthew Anakin
date: "CG3CG3"
output: "html_document"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# 1. Clear environment (optional but recommended to prevent ghost variables)
rm(list = ls())

# 2. Set seed and define n
set.seed(33)
n <- 100

# 3. Generate the dataframe
data <- data.frame(
  StudyHours = round(runif(n, 2, 20), 1),
  AttendanceRate = round(runif(n, 60, 100), 1),
  Gender = factor(sample(c("Male", "Female"), n, replace = TRUE)),
  SchoolType = factor(sample(c("Public", "Private"), n, replace = TRUE))
)

# 4. Calculate ScienceScore
data$ScienceScore <- round(
  40 +
    2.5 * data$StudyHours +
    0.3 * data$AttendanceRate +
    ifelse(data$Gender == "Female", 3, 0) +
    ifelse(data$SchoolType == "Private", 5, 0) +
    rnorm(n, 0, 5), 1
)

head(data)
model <- lm(ScienceScore ~ ., data = data)

# View the coefficients (Intercept and Slope) to form your equation
coef(model)

summary(model)
```

The regression model is statistically significant, as indicated by the F-statistic (F = 153.8, p < 2.2 × 10⁻¹⁶). This suggests that the set of independent variables—StudyHours, AttendanceRate, Gender, and SchoolType—collectively have a significant effect on ScienceScore. Therefore, the model is appropriate for explaining variations in the dependent variable.

# Goodness of the Fit
The model achieved a Multiple R-squared of 0.8662 and an Adjusted R-squared of 0.8606, indicating that approximately 86% of the variability in ScienceScore is explained by the predictors included in the model. This reflects a strong fit and suggests that the model has high explanatory power.

# Residuals Analysis

The residuals have the following distribution:
- Minimum: -15.79

- 1st Quartile: -3.49

- Median: 0.30

- 3rd Quartile: 3.51

- Maximum: 11.80

The median residual is close to zero, indicating that the model does not exhibit systematic bias in prediction. The spread of residuals appears reasonably symmetric, suggesting that the assumption of linearity and unbiased errors is satisfied. The Residual Standard Error (5.26) indicates that the average deviation of predicted values from actual values is approximately ±5.26 units.

# Significance of Coefficients
- StudyHours (β = 2.4586, p < 2e-16): Highly significant predictor, indicating a strong positive relationship with ScienceScore.

- AttendanceRate (β = 0.2900, p = 4.55e-09): Statistically significant, with a positive contribution to the model.

- SchoolTypePublic (β = -4.5174, p = 5.29e-05): Significant negative effect, suggesting lower scores for public school students relative to private school students.

- GenderMale (β = -2.1034, p = 0.0508): Marginally significant at the 5% level, indicating a weak negative effect. This variable may require further investigation or validation.

# Normality

## QQplot

```{r}
qqnorm (residuals(model), ylab="Residuals")
qqline(residuals(model))
```

- The Q-Q plot shows that your residuals follow the normal distribution well in the middle (the bulk of your data), as points closely track the diagonal line. However, both tails deviate from the line — particularly the lower tail, where a few extreme negative residuals (around −16) pull away noticeably. This indicates mild heavy-tailed non-normality, driven largely by a small number of outliers rather than a systematic distributional problem.

## Shapiro Test

```{r}
length(residuals(model))
shapiro.test(residuals(model))
```

-   Does not reject the null hypothesis

## Statistical Test

```{r}
resid_std <- scale(residuals(model)) # mean=0, sd=1
ks.test(resid_std, "pnorm")
```

# Homoscedasticity

## Graphical Analysis of the residuals vs fitted values

```{r}
plot(fitted(model), residuals(model))
```

-  The residuals vs. fitted plot shows a random scatter around zero with no clear curve or systematic pattern, suggesting that both the linearity assumption and homoscedasticity are reasonably satisfied across the range of fitted values (70–120). However, a couple of extreme negative residuals (around −16) stand out as potential outliers that warrant further investigation.

## Statistical Test

```{r}
library("lmtest")
bptest(model)
```

-   Homoscedasticity is not present in the model since the p-value is greater than 0.05

## Goldfeld Quandt Test

```{r}
gqtest(model, order.by = ScienceScore ~ ., data = data)
```

-   GQ value \> 1 suggests variance might be larger in the second segment

-   p-value = 0.0611 which is greater than 0.05 suggests that there is no heteroscedasticity which also makes it borderline significant

# Multicollinearity

PURPOSE OF MULTICOLLINEARITY: Check if the features are not highly correlated with each other

```{r}
options(width = 50)
library(car)
vif(model)
# VIF > 5 -> moderate concern
# VIF > 10 -> serious multicollinearity
```

Since there is no high correlation with other independent variables, so it does not have a multicollinearity issue.

# Independence Test

## Graphical Analysis of Residuals

```{r}
n = length(residuals(model))
plot(tail(residuals(model), n-1) ~
head(residuals(model), n-1),
xlab=expression(hat(epsilon)[i]),
ylab=expression(hat(epsilon)[i+1]))
abline(h=0, v=0, col = grey(0.5))
```

- This plot displays the current residuals (ε̂ᵢ) against the lagged residuals (ε̂ᵢ₋₁) to visually check for autocorrelation among the residuals. The scatter appears random with no clear directional pattern or clustering in any quadrant, suggesting that consecutive residuals are not systematically related to one another. This indicates that the assumption of independence of errors is likely satisfied, meaning there is no significant autocorrelation present in the model.

## Statistical Test

```{r}
dwtest(model)
```

-   The p-value indicates no evidence of serial correlation. Therefore, we can assume that the errors are uncorrelated

# Regression Diagnostics Plot

```{r}
par(mfrow = c(2, 2))
plot(model)
```

-   Here are the diagnostic plots for each regression

The Residuals vs Fitted and Scale-Location plots show no strong systematic pattern, supporting the assumptions of linearity and homoscedasticity, though the slight downward trend in the Scale-Location plot suggests a minor decrease in variance at higher fitted values. The Q-Q plot confirms approximate normality for most residuals with slight heavy tails, while the Residuals vs Leverage plot shows that the flagged observations (points 14, 15) have moderate leverage but appear to remain within acceptable Cook's distance boundaries, meaning they are unlikely to be severely distorting the model estimates. Overall, the diagnostics suggest that the model assumptions are reasonably met, with only minor deviations that do not appear to significantly impact the validity of the regression results.



# Additional Diagnostics Plots

```{r}


#diagnostic stuff


# study hours


par(mfrow = c(2, 2))


plot(data$StudyHours, residuals(model),
     xlab = "Study Hours", ylab = "Residuals",
     main = "Residuals vs Study Hours")
abline(h = 0, col = "red")



#attendance

plot(data$AttendanceRate, residuals(model),
     xlab = "Attendance Rate", ylab = "Residuals",
     main = "Residuals vs Attendance Rate")
abline(h = 0, col = "red")


#gender

boxplot(residuals(model) ~ data$Gender,
        xlab = "Gender", ylab = "Residuals",
        main = "Residuals by Gender")
abline(h = 0, col = "red")

#school type

boxplot(residuals(model) ~ data$SchoolType,
        xlab = "School Type", ylab = "Residuals",
        main = "Residuals by School Type")
abline(h = 0, col = "red")

par(mfrow = c(1, 1))  #so it turns to a grid 

```
