---
title: "CG3CG3"
author: 
- de Pedro, Joenell Conrad
- Florentino, Ian Miguel
- Nacu, Drixyl Reece
- Ruelan, Ivan Paul
- Seno, Matthew Anakin
date: "CG3CG3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(arules)
library(arulesViz)
```

# 1. Load the dataset and convert it to a transaction object using arules
```{r}
setwd("C:/stuff/Code/R/New folder/g3-data-analysis/src") # sets ur path here
data <- read.csv("./OnlineRetail.csv")
head(data)
```

```{r}
transactions <- as(split(data$Description, data$InvoiceNo), "transactions")

summary(transactions)
```
# 2. Use the Apriori algorithm to find frequent itemsets with minimum support and confidence.
```{r}
rules <- apriori(
  transactions,
  parameter = list(
    supp = 0.02,
    conf = 0.5,
    minlen = 2
  )
)
```

# 3. Inspect the top rules sorted by lift.

```{r}
print("Top Rules sorted by lift")
options(width = 40)
rules_sorted <- sort(rules, by="lift", decreasing=TRUE)
inspect(rules_sorted[1:10])
# inspect(head(rules_sorted, 10))
```
# 4. Visualize the rules using arulesViz.
```{r}
plot(rules, method="scatterplot",
     measure=c("support","confidence"),
     shading="confidence")
```
```{r}
plot(rules, method="graph")
```

# 5. Discuss practical business implications (e.g., product bundling, promotions).
- Common pairs can be found in the frequently sold products meaning we can sell SE- SETX :))
- pag add mo

# Discussion

## Which items are frequently purchased together? (answer comes from question 3)
- GREEN REGENCY TEACUP AND SAUCER and PINK REGENCY TEACUP AND SAUCER and ROSES REGENCY TEACUP AND SAUCER => 89% confidence
- GARDENERS KNEELING PAD CUP OF TEA and GARDENERS KNEELING PAD KEEP CALM =? 71%
- SPACEBOY LUNCH BOX and DOLLY GIRL LUNCH BOX  => 59% confidence


## How could the store use this information to increase sales?
- Could offer product bundling or buy multiple get percentage off offers
- add lang mu
## Are there rules that suggest cross-selling opportunities?
- Yes, teacup and saucer color sets, lunch box sets

## What is the effect of changing support or confidence on the number of rules?
- kamo ani hehe