---
title: "CG3CG3"
author: 
- de Pedro, Joenell Conrad
- Florentino, Ian Miguel
- Nacu, Drixyl Reece
- Ruelan, Ivan Paul
- Seno, Matthew Anakin
date: "CG3CG3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(arules)
library(arulesViz)
```

# 1. Load the dataset and convert it to a transaction object using arules
```{r}
setwd("C:/Nyeh/Code/R/g3-data-analysis/src")
data <- read.csv("./OnlineRetail.csv")
head(data)
```

```{r}
transactions <- as(split(data$Description, data$InvoiceNo), "transactions")

summary(transactions)
```
# 2. Use the Apriori algorithm to find frequent itemsets with minimum support and confidence.
```{r}
rules <- apriori(
  transactions,
  parameter = list(
    supp = 0.02,
    conf = 0.5,
    minlen = 2
  )
)
```

# 3. Inspect the top rules sorted by lift.

```{r}
print("Top Rules sorted by lift")
options(width = 40)
rules_sorted <- sort(rules, by="lift", decreasing=TRUE)
inspect(rules_sorted[1:10])
# inspect(head(rules_sorted, 10))
```
# 4. Visualize the rules using arulesViz.
```{r}
plot(rules, method="scatterplot",
     measure=c("support","confidence"),
     shading="confidence")
```
```{r}
plot(rules, method="graph")
```

# 5. Discuss practical business implications (e.g., product bundling, promotions).
- Common pairs can be found in the frequently sold products meaning we can sell SE- SETX :))

# Discussion

## Which items are frequently purchased together?
[1] {GREEN REGENCY TEACUP AND SAUCER,
     ROSES REGENCY TEACUP AND SAUCER }   => {PINK REGENCY TEACUP AND SAUCER}     0.02119691  0.7002551 0.03027027 22.64246   549
[2] {PINK REGENCY TEACUP AND SAUCER,
     ROSES REGENCY TEACUP AND SAUCER }   => {GREEN REGENCY TEACUP AND SAUCER}    0.02119691  0.8941368 0.02370656 21.90931   549
[3] {GARDENERS KNEELING PAD CUP OF TEA } => {GARDENERS KNEELING PAD KEEP CALM }  0.02119691  0.7176471 0.02953668 20.11586   549
[4] {GARDENERS KNEELING PAD KEEP CALM }  => {GARDENERS KNEELING PAD CUP OF TEA } 0.02119691  0.5941558 0.03567568 20.11586   549
[5] {GREEN REGENCY TEACUP AND SAUCER,
     PINK REGENCY TEACUP AND SAUCER}     => {ROSES REGENCY TEACUP AND SAUCER }   0.02119691  0.8524845 0.02486486 19.71370   549   

## How could the store use this information to increase sales?
- Could offer product bundling or buy multiple get percentage off offers
- add lang mu
## Are there rules that suggest cross-selling opportunities?
- Yes,teacup and saucer color sets, lunch box sets

## What is the effect of changing support or confidence on the number of rules?
- kamo ani hehe